---
title: "Income Inequality and Labour Market Outcomes in Canada 2006-2024"
author: "Alex McKeown"
date: "2025-03-29"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    fig_caption: true
    number_sections: true
    theme: cosmo
    css: styles.css
    highlight: tango
    code_folding: hide
    self_contained: true
    df_print: paged
    mathjax: default
---

```{css, echo=FALSE}
body {
  font-family: "Times New Roman", Times, serif;
  line-height: 2;
  font-size: 12pt;
}

h1, h2, h3, h4, h5, h6 {
  font-family: "Times New Roman", Times, serif;
  font-weight: bold;
}

.main-container {
  max-width: 8.5in;
  margin: auto;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,       # Changed from FALSE to TRUE to allow code folding
  warning = FALSE,    
  message = FALSE,    
  fig.width = 10,     
  fig.height = 6,     
  fig.align = "center", 
  dpi = 300,
  gc.verbose = FALSE  # Suppress garbage collection output
)

# Make sure the required packages for code folding are loaded
if (!requireNamespace("rmarkdown", quietly = TRUE)) {
  install.packages("rmarkdown")
}
if (!requireNamespace("knitr", quietly = TRUE)) {
  install.packages("knitr")
}
```

```{r packages-and-data, include=FALSE}
# Setup - Clear workspace and manage packages
# Define required packages, install if needed, load all packages, set working directory,
# load and combine data from CSV files matching the pattern "^pub(01|06)\\d{2}\\.csv$"
rm(list = ls())
invisible(gc())

required_packages <- c(
  "data.table", "dplyr", "tidyverse", "plm",
  "caret", "cluster", "modelsummary", "ineq", "ggplot2", "broom", 
  "rmarkdown"
)

packages_loaded <- sapply(required_packages, function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    message(paste("Installing package:", pkg))
    tryCatch({
        install.packages(pkg, dependencies = TRUE)
        library(pkg, character.only = TRUE)
        return(TRUE)
      }, error = function(e) {
        message(paste("Failed to install package:", pkg, "Error:", e$message))
        return(FALSE)
      })
  } else {
    library(pkg, character.only = TRUE)
    return(TRUE)
  }
})

if (!packages_loaded["data.table"]) {
  stop("Critical package 'data.table' could not be loaded. Aborting.")
}

csv_folder_path <- "/Users/alex/Desktop/ECO340-Project" # USER: MODIFY THIS PATH
tryCatch({
    setwd(csv_folder_path)
  }, warning = function(w){
    stop("Warning setting working directory: ", w$message, ". Check path: ", csv_folder_path)
  }, error = function(e) {
    stop("Error setting working directory: ", e$message, ". Check path: ", csv_folder_path)
  }
)

# Find and load CSV files matching the pattern
all_csv_files <- list.files(path = getwd(), pattern = "\\.csv$", full.names = TRUE, ignore.case = TRUE)
filename_pattern <- "^pub(01|06)\\d{2}\\.csv$"
all_basenames <- basename(all_csv_files)
matching_indices <- grepl(filename_pattern, all_basenames, ignore.case = TRUE)
csv_files <- all_csv_files[matching_indices]

if (length(csv_files) == 0) {
  stop("No CSV files matching the pattern '", filename_pattern, "' were found in ", getwd())
} else {
  LFS_data <- NULL
  for (i in seq_along(csv_files)) {
    file_path <- csv_files[i]

    current_data <- tryCatch({
        data.table::fread(file_path, showProgress = FALSE)
      }, error = function(e) {
        return(NULL)
      })

    if (!is.null(current_data) && nrow(current_data) > 0) {
      if (is.null(LFS_data)) {
        LFS_data <- current_data
      } else {
        LFS_data <- data.table::rbindlist(list(LFS_data, current_data), use.names = TRUE, fill = TRUE)
      }
    }
    rm(current_data)
    invisible(gc())
  }
  rm(all_csv_files, all_basenames, matching_indices, csv_files, file_path)
  invisible(gc())
}
```

# Introduction

This analysis hopes to help better unpack income inequality in the Canadian context. This analysis investigates trends in income inequality within different groupings of Labour Force Survey (LFS) observations over a near eighteen year period, from January 2006 to June 2024 – every 6th month of LFS data was used for this analysis, to put less stress on memory constraints. Using this data, a multivariable regression model to predict hourly earnings based on key variables was built. Then, using this model to identify potentially interesting areas of difference, subsets of the LFS survey sample were constructed, creating a time series of Gini coefficients for hourly income within each group selected. For instance, comparing trends over time for the Gini coefficient within the male population versus the female population. It is important to note that this report uses the Public Use Microdata File (PUMF) which contains less information than the full data – the PUMF omits information such as race, city/municipality located, etc. As well, the LFS is a repeated cross-section of the population, it is not a cohort study like the National Longitudinal Survey of Youth (NLSY) conducted by the US Bureau of Labour Statistics, so techniques applicable to that dataset, such as a fixed effect model, are not applicable here.

# Data Cleaning and Variable Creation

```{r data-cleaning, include=FALSE}
# Data Cleaning & Variable Creation
# Convert variables to appropriate formats and create binary indicators for analysis

setDT(LFS_data)

#Log Hourly Earnings
LFS_data[, ln_income := log(HRLYEARN)]

# Employment and Labour Force Status
LFS_data[, is_employed := fifelse(LFSSTAT == 1, 1, 0)]
LFS_data[, in_lf := fifelse(LFSSTAT %in% c(1, 2), 1, 0)]

# Sex
LFS_data[, woman := fifelse(SEX == 2 | GENDER == 2, 1L, 0L, na = 0L)]

# Age
LFS_data[, rough_age_dock15 := fcase(
    AGE_12 == 1,  2, AGE_12 == 2,  7, AGE_12 == 3,  12, AGE_12 == 4,  17,
    AGE_12 == 5,  22, AGE_12 == 6,  27, AGE_12 == 7,  32, AGE_12 == 8,  37,
    AGE_12 == 9,  42, AGE_12 == 10, 47, AGE_12 == 11, 52, AGE_12 == 12, 57,
    default = NA_real_
)]

# Province indicators
LFS_data[, prov_nl := fifelse(PROV == 10, 1, 0)]
LFS_data[, prov_pe := fifelse(PROV == 11, 1, 0)]
LFS_data[, prov_ns := fifelse(PROV == 12, 1, 0)]
LFS_data[, prov_nb := fifelse(PROV == 13, 1, 0)]
LFS_data[, prov_qc := fifelse(PROV == 24, 1, 0)]
LFS_data[, prov_mb := fifelse(PROV == 46, 1, 0)]
LFS_data[, prov_sk := fifelse(PROV == 47, 1, 0)]
LFS_data[, prov_ab := fifelse(PROV == 48, 1, 0)]
LFS_data[, prov_bc := fifelse(PROV == 59, 1, 0)]
province_dummies <- c("prov_nl", "prov_pe", "prov_ns", "prov_nb", "prov_qc", "prov_mb", "prov_sk", "prov_ab", "prov_bc")

# Union status
LFS_data[, union_member := fifelse(UNION == 1, 1, 0)]
LFS_data[, union_contract := fifelse(UNION %in% c(1, 2), 1, 0)]

# Immigration status
LFS_data[, newer_immigrant := fifelse(IMMIG == 1, 1, 0)]
LFS_data[, older_immigrant := fifelse(IMMIG == 2, 1, 0)]

# Education dummies

LFS_data[, zero_to_eight := fifelse(EDUC == 0, 1, 0)]
LFS_data[, some_hs := fifelse(EDUC == 1, 1, 0)]
LFS_data[, some_ps := fifelse(EDUC == 3, 1, 0)]
LFS_data[, ps_cert := fifelse(EDUC == 4, 1, 0)]
LFS_data[, ba := fifelse(EDUC == 5, 1, 0)]
LFS_data[, above_ba := fifelse(EDUC == 6, 1, 0)]
education_dummies <- c("zero_to_eight", "some_hs", "some_ps", "ps_cert", "ba", "above_ba")

# NAICS industry dummies
LFS_data[, naics_ForestryLogSupport := fifelse(NAICS_21 == 2, 1, 0)]
LFS_data[, naics_FishingHuntTrap := fifelse(NAICS_21 == 3, 1, 0)]
LFS_data[, naics_MiningOilGasExtract := fifelse(NAICS_21 == 4, 1, 0)]
LFS_data[, naics_Utilities := fifelse(NAICS_21 == 5, 1, 0)]
LFS_data[, naics_Construction := fifelse(NAICS_21 == 6, 1, 0)]
LFS_data[, naics_MfgDurable := fifelse(NAICS_21 == 7, 1, 0)]
LFS_data[, naics_MfgNonDurable := fifelse(NAICS_21 == 8, 1, 0)]
LFS_data[, naics_Wholesale := fifelse(NAICS_21 == 9, 1, 0)]
LFS_data[, naics_Retail := fifelse(NAICS_21 == 10, 1, 0)]
LFS_data[, naics_TransportWarehousing := fifelse(NAICS_21 == 11, 1, 0)]
LFS_data[, naics_FinanceInsurance := fifelse(NAICS_21 == 12, 1, 0)]
LFS_data[, naics_RealEstateRentalLeasing := fifelse(NAICS_21 == 13, 1, 0)]
LFS_data[, naics_ProfSciTech := fifelse(NAICS_21 == 14, 1, 0)]
LFS_data[, naics_BusinessBuildingSupport := fifelse(NAICS_21 == 15, 1, 0)]
LFS_data[, naics_Educational := fifelse(NAICS_21 == 16, 1, 0)]
LFS_data[, naics_HealthSocialAssist := fifelse(NAICS_21 == 17, 1, 0)]
LFS_data[, naics_InfoCultRec := fifelse(NAICS_21 == 18, 1, 0)]
LFS_data[, naics_AccomFood := fifelse(NAICS_21 == 19, 1, 0)]
LFS_data[, naics_OtherServices := fifelse(NAICS_21 == 20, 1, 0)]
LFS_data[, naics_PublicAdmin := fifelse(NAICS_21 == 21, 1, 0)]
naics_dummies <- c(
  "naics_ForestryLogSupport", "naics_FishingHuntTrap", "naics_MiningOilGasExtract", "naics_Utilities",
  "naics_Construction", "naics_MfgDurable", "naics_MfgNonDurable", "naics_Wholesale",
  "naics_Retail", "naics_TransportWarehousing", "naics_FinanceInsurance", "naics_RealEstateRentalLeasing",
  "naics_ProfSciTech", "naics_BusinessBuildingSupport", "naics_Educational", "naics_HealthSocialAssist",
  "naics_InfoCultRec", "naics_AccomFood", "naics_OtherServices", "naics_PublicAdmin"
)

# Time variables
LFS_data[, years_from_start := (SURVYEAR - 2006) + ifelse(SURVMNTH == 6, 0.5, 0)]
yfs_dummy_names <- character(0)
unique_vals <- sort(unique(na.omit(LFS_data$years_from_start)))

if (length(unique_vals) > 0) {
  for (val in unique_vals) {
    new_col_name <- paste0("yfs_", gsub("[^a-zA-Z0-9_]", "_", as.character(val)))
    yfs_dummy_names <- c(yfs_dummy_names, new_col_name)
    LFS_data[, (new_col_name) := fifelse(years_from_start == val, 1L, 0L, na = 0L)]
  }
}

# Categorical variables for grouping
LFS_data[, province_name := fcase(
    PROV == 10, "Newfoundland and Labrador", PROV == 11, "Prince Edward Island",
    PROV == 12, "Nova Scotia", PROV == 13, "New Brunswick",
    PROV == 24, "Quebec", PROV == 35, "Ontario",
    PROV == 46, "Manitoba", PROV == 47, "Saskatchewan",
    PROV == 48, "Alberta", PROV == 59, "British Columbia"
)]

LFS_data[, named_sex := fcase(woman == 0, "Male", woman == 1, "Female")]

LFS_data[, age_group := fcase(
    AGE_12 == 1, "15_19",
    AGE_12 %in% c(2, 3, 4), "20_34",
    AGE_12 %in% c(5, 6, 7, 8), "35_54",
    AGE_12 %in% c(9, 10, 11, 12), "55+",
    default = NA_character_
)]

rm(unique_vals, val, new_col_name)
invisible(gc())
```

To make analysis easier, I created several derived variables from the raw LFS data. Variables with multiple categories in numeric form, such as province codes, were converted to binary dummy variables. The key transformations include:

-   Gender: re-coded from 1 & 2 to 0 & 1 for regression analysis
-   Union status: created separate indicators for union membership and being covered by a union contract
-   Provinces: created 9 dummy variables with Ontario as the reference province
-   Education: created dummy variables, one for each level of education save the reference, which I set to high school grad rather than 0-8 years as I feel it is more intuitive when interpreting regression results – we don’t care as much about how college impacts earnings relative to someone who didn’t complete highschool.
-   Age: approximated using AGE12 variable, creating a running variable and labels for subsetting later on
-   Labour force status: converted LFSSTAT into two dummies for employment and labour force participation
-   Immigration status: created dummy variables for recent immigrants (≤10 years) and established immigrants
-   Income: created a logarithmic version of hourly earnings for easier interpretation in regressions

The data.table package was used rather than the mutate function. The reason for this is because initially, I was stubbornly trying to do this analysis using every month of LFS survey data from 2006 through to the most recent release. In hindsight, that was a really silly decision, and I should have done this approach of looking at every few waves from the get go. Alternatively, I could have taken a sample of each sample, though this may wind up introducing different biases into the data, if say certain smaller provinces were under-represented (alternatively alternatively I could have swapped from working on my dated MacBook to a more powerful machine with 8x the RAM, but keeping myself constrained both a) encourages sensible code & optimization for memory and b) makes sure that whoever reads this can run the code on their machine).



# Wage Determinants: Regression Analysis

This section examines the impact of key variables and controls on hourly earnings. As this model is estimated by applying it to multiple waves of LFS data, year dummy variables are included to account for time fixed-effects and year-specific events. The analysis is restricted to employed Canadians, with variations in unemployment and labour force participation covered later.

The purpose of this section is not to derive any profound insights into the determinants of wages in Canada, rather the goal is to create a starting point for how we can divide up our sample to compare the trends of Gini coefficents over time.

The dependent variable in these regressions is the natural logarithm of hourly earnings, for easier interpretation of coefficients. The "base" or intercept observation is a male high school graduate from Ontario who is not in a union, nor is covered by a union contract, nor is an immigrant, and works in the Agriculture sector.

```{r wage-regressions}
# Store the models for comparison
models <- list()

# Model 1: Basic with gender and year effects only
dummy_formula_part <- paste(yfs_dummy_names, collapse = " + ")
regression_formula <- as.formula(paste("ln_income ~ woman +", dummy_formula_part))
models[[1]] <- lm(regression_formula, data = LFS_data)
rm(regression_formula, dummy_formula_part)
invisible(gc())

# Model 2: Adding province effects
dummy_formula_part1 <- paste(yfs_dummy_names, collapse = " + ")
dummy_formula_part2 <- paste(province_dummies, collapse = " + ")
regression_formula <- as.formula(paste("ln_income ~ woman + ", dummy_formula_part1, " + ", dummy_formula_part2))
models[[2]] <- lm(regression_formula, data = LFS_data)
rm(regression_formula, dummy_formula_part1, dummy_formula_part2)
invisible(gc())

# Model 3: Adding education effects
dummy_formula_part1 <- paste(yfs_dummy_names, collapse = " + ")
dummy_formula_part2 <- paste(province_dummies, collapse = " + ")
dummy_formula_part3 <- paste(education_dummies, collapse = " + ")
regression_formula <- as.formula(paste("ln_income ~ woman + ", dummy_formula_part1, " + ", dummy_formula_part2, " + ", dummy_formula_part3))
models[[3]] <- lm(regression_formula, data = LFS_data)
rm(regression_formula, dummy_formula_part1, dummy_formula_part2, dummy_formula_part3)
invisible(gc())

# Model 4: Adding age polynomial
dummy_formula_part1 <- paste(yfs_dummy_names, collapse = " + ")
dummy_formula_part2 <- paste(province_dummies, collapse = " + ")
dummy_formula_part3 <- paste(education_dummies, collapse = " + ")
regression_formula <- as.formula(paste("ln_income ~ woman + poly(rough_age_dock15, 2, raw=TRUE) + ",
                                     dummy_formula_part1, " + ", dummy_formula_part2, " + ", dummy_formula_part3))
models[[4]] <- lm(regression_formula, data = LFS_data)
rm(regression_formula, dummy_formula_part1, dummy_formula_part2, dummy_formula_part3)
invisible(gc())

# Model 5: Full model with union, immigration, and industry sectors
dummy_formula_part1 <- paste(yfs_dummy_names, collapse = " + ")
dummy_formula_part2 <- paste(province_dummies, collapse = " + ")
dummy_formula_part3 <- paste(education_dummies, collapse = " + ")
dummy_formula_part4 <- paste(naics_dummies, collapse = " + ")
regression_formula <- as.formula(paste("ln_income ~ woman + poly(rough_age_dock15, 2, raw=TRUE) + newer_immigrant + older_immigrant + union_contract + ",
                                     dummy_formula_part1, " + ", dummy_formula_part2, " + ", dummy_formula_part3, " + ", dummy_formula_part4))
models[[5]] <- lm(regression_formula, data = LFS_data)
rm(regression_formula, dummy_formula_part1, dummy_formula_part2, dummy_formula_part3, dummy_formula_part4)
invisible(gc())

# Define model names for the table
model_names <- c("Gender Only", "Add Provinces", "Add Education", "Add Age", "Full Model")
names(models) <- model_names

# Create a custom coefficient map for the table
coef_map <- c(
  "(Intercept)" = "Intercept",
  "woman" = "Female",
  "poly(rough_age_dock15, 2, raw = TRUE)1" = "Age",
  "poly(rough_age_dock15, 2, raw = TRUE)2" = "Age²",
  "newer_immigrant" = "Recent Immigrant (≤10yrs)",
  "older_immigrant" = "Established Immigrant (>10yrs)",
  "union_contract" = "Union Coverage",
  
  # Key education variables
  "zero_to_eight" = "Education: 0-8 years",
  "ba" = "Education: Bachelor's degree",
  "above_ba" = "Education: Above bachelor's",
  
  # Selected provinces
  "prov_qc" = "Quebec",
  "prov_ab" = "Alberta",
  "prov_bc" = "British Columbia",
  
  # Selected industry sectors
  "naics_MiningOilGasExtract" = "Industry: Mining & Oil Extraction",
  "naics_FinanceInsurance" = "Industry: Finance & Insurance",
  "naics_HealthSocialAssist" = "Industry: Healthcare & Social Assist."
)

# Generate regression table
modelsummary(models,
             title = "Log Wage Regressions with Different Specifications",
             stars = TRUE,
             coef_map = coef_map,
             gof_map = c("nobs", "r.squared", "adj.r.squared"),
             notes = "Year fixed effects included in all models. Ontario is the reference province. Full range of dummy variables not shown for ease of viewing",
             output = "html")


```

```{r all-models-full-comparison-collapsible, echo=FALSE, results='asis'}
## Comparison of All Regression Models (Collapsible)
# This chunk generates a collapsible HTML table comparing all regression models
# and displaying all coefficients.
# The code itself is hidden (echo=FALSE).
# results='asis' allows us to print raw HTML tags directly.

# Ensure the models list exists and contains models
if (!exists("models") || length(models) == 0) {
  stop("Regression models not found. Please ensure the 'wage-regressions' chunk ran successfully.")
}

# Start the collapsible section using HTML <details> tag
cat('<details>')
# Add a summary line that the user clicks to expand/collapse
cat('<summary>Click to show/hide the full comparison table of all models</summary>')

# Generate the modelsummary comparison table for ALL models.
# Omit coef_map to show all coefficients.
# With output='html' and results='asis', modelsummary prints the
# HTML table code directly into the document at this point.
modelsummary(models, # Pass the list of all models
             title = "Comparison of All Log Wage Regression Models (Full Coefficients)",
             stars = TRUE,
             gof_map = c("nobs", "r.squared", "adj.r.squared"),
             notes = "Includes all dummy variables (Year, Province, Education, Industry) where included in the model. Reference categories: Ontario, High School Graduate, Agriculture.",
             output = "html")

# End the collapsible section with the closing </details> tag
cat('</details>')
```

``` {r}
# Clear models to free memory
rm(models, model_names, coef_map)
invisible(gc())
```

## Interpretation of Results

The regression results above show how different factors influence hourly wages in Canada. Key observations include:

1.  **Gender Wage Gap**: Across all specifications, women earn less than men, though the gap changes a bit depending on what controls are added. As this gap is persistent, for our later analysis of Gini coefficients between groups, it seems wise to create different groups for women and men.

2.  **Provincial Variation**: Substantial differences exist across provinces, with Alberta showing a wage premium and Quebec showing somewhat lower wages compared to Ontario (the reference province). While the Federal government through income taxes and other means of taxation has a large role to play in combating inequality, it seems pertinent to divide by provinces as well, as there are differences in provincial services, cost of living, sectors, etc.

3.  **Education Premium**: Higher education levels are associated with substantially higher wages, with the greatest returns to Bachelor's and advanced degrees. This seems like a source of inequity between workers which we do want to capture.

4.  **Age Effects**: Wages follow a concave pattern with age, increasing at a decreasing rate and eventually plateauing. If inequalities are caused by life cycles, young workers make less than older workers for example, that's not too problematic to most, young or old, because the young workers will one day be those older workers. Thus in our analysis of Gini coefficients, we will be dividing into age groups.

5.  **Immigration Status**: Recent immigrants face a wage penalty that persists even after controlling for education and industry. This may be due a variety of factors, we will examine this later on.

6.  **Union Coverage**: Union contracts are associated with higher wages, reflecting the collective bargaining premium. This does not seem like an important distinction to make in analysis of Gini coefficients -- if say, union membership is on the decline, but unions are still powerful and give the fewer workers in them a much greater wage, we would want to see that in data for Gini coefficients of say 25-34 year olds.

7.  **Industry Effects**: Significant wage differences exist across industries, with Mining & Oil Extraction and Finance & Insurance offering premiums compared to the Agriculture sector (the reference industry). For the same reason given as with education premiums, this is something which should be included when looking at Gini coefficients for certain segments.

# Income Inequality Analysis

This section examines trends in income inequality across different demographic groups in Canada. Using the Gini coefficient as a measure of inequality, I analyze how wage dispersion has evolved over time for each gender, province, immigration status, and education level subgroup.

```{r gini-calculation}
# Function to calculate Gini coefficient for each subset by year
calculate_gini_by_year <- function(data_subset) {
  if (is.null(data_subset) || nrow(data_subset) < 10) {
    return(data.frame(years_from_start = numeric(0), gini = numeric(0), median_income = numeric(0)))
  }
  
  result <- data_subset[!is.na(HRLYEARN), .(
    gini = as.numeric(ineq::ineq(HRLYEARN, type="Gini")),
    median_income = as.numeric(median(HRLYEARN, na.rm = TRUE)),
    count = .N
  ), by = years_from_start]
  
  result <- result[count >= 30]
  return(result[, .(years_from_start, gini, median_income)])
}

# Process gender subsets first
gender_subsets <- split(LFS_data, LFS_data$named_sex)
gender_gini_results <- data.frame(
  group = character(),
  years_from_start = numeric(),
  gini = numeric(),
  median_income = numeric(),
  stringsAsFactors = FALSE
)

for (gender_name in names(gender_subsets)) {
  subset_gini <- calculate_gini_by_year(gender_subsets[[gender_name]])
  
  if (nrow(subset_gini) > 0) {
    subset_gini_df <- data.frame(
      group = gender_name,
      years_from_start = subset_gini$years_from_start,
      gini = subset_gini$gini,
      median_income = subset_gini$median_income,
      stringsAsFactors = FALSE
    )
    gender_gini_results <- rbind(gender_gini_results, subset_gini_df)
  }
}

# Convert to actual years and create gender plot
gender_gini_results$year <- 2006 + gender_gini_results$years_from_start
p <- ggplot(gender_gini_results, aes(x = year, y = gini, color = group)) +
  geom_line(linewidth = 1.2) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
  labs(title = "Income Inequality (Gini) Trends by Gender",
       subtitle = "All of Canada",
       x = "Year",
       y = "Gini Coefficient",
       color = "Gender") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(limits = c(0.2, 0.35))
print(p)
rm(p, gender_subsets, gender_gini_results, subset_gini, subset_gini_df, gender_name)
invisible(gc())

# Process province subsets
province_subsets <- split(LFS_data, LFS_data$province_name)
province_gini_results <- data.frame(
  group = character(),
  years_from_start = numeric(),
  gini = numeric(),
  median_income = numeric(),
  stringsAsFactors = FALSE
)

for (province_name in names(province_subsets)) {
  subset_gini <- calculate_gini_by_year(province_subsets[[province_name]])
  
  if (nrow(subset_gini) > 0) {
    subset_gini_df <- data.frame(
      group = province_name,
      years_from_start = subset_gini$years_from_start,
      gini = subset_gini$gini,
      median_income = subset_gini$median_income,
      stringsAsFactors = FALSE
    )
    province_gini_results <- rbind(province_gini_results, subset_gini_df)
  }
}

# Convert to actual years and create province plot
province_gini_results$year <- 2006 + province_gini_results$years_from_start
p <- ggplot(province_gini_results, aes(x = year, y = gini, color = group)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, alpha = 0.2) +
  labs(title = "Income Inequality Trends by Province",
       x = "Year",
       y = "Gini Coefficient",
       color = "Province") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 2)) +
  scale_y_continuous(limits = c(0.2, 0.35))
print(p)
rm(p, province_subsets, province_gini_results, subset_gini, subset_gini_df, province_name)
invisible(gc())

# Process immigration status subsets
LFS_data[, immigration_status := fcase(
  newer_immigrant == 1, "Recent Immigrant",
  older_immigrant == 1, "Established Immigrant",
  default = "Canadian-born"
)]
immigration_subsets <- split(LFS_data, LFS_data$immigration_status)
immigration_gini_results <- data.frame(
  group = character(),
  years_from_start = numeric(),
  gini = numeric(),
  median_income = numeric(),
  stringsAsFactors = FALSE
)

for (immigration_name in names(immigration_subsets)) {
  subset_gini <- calculate_gini_by_year(immigration_subsets[[immigration_name]])
  
  if (nrow(subset_gini) > 0) {
    subset_gini_df <- data.frame(
      group = immigration_name,
      years_from_start = subset_gini$years_from_start,
      gini = subset_gini$gini,
      median_income = subset_gini$median_income,
      stringsAsFactors = FALSE
    )
    immigration_gini_results <- rbind(immigration_gini_results, subset_gini_df)
  }
}

# Convert to actual years and create immigration plot
immigration_gini_results$year <- 2006 + immigration_gini_results$years_from_start
p <- ggplot(immigration_gini_results, aes(x = year, y = gini, color = group)) +
  geom_line(linewidth = 1.2) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
  labs(title = "Income Inequality Trends by Immigration Status",
       x = "Year",
       y = "Gini Coefficient",
       color = "Immigration Status") +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2") +
  scale_y_continuous(limits = c(0.2, 0.35))
print(p)
rm(p, immigration_subsets, immigration_gini_results, subset_gini, subset_gini_df, immigration_name)
invisible(gc())

# Process education level subsets
LFS_data[, education_level := fcase(
  zero_to_eight == 1, "0-8 years",
  some_hs == 1, "Some high school",
  EDUC == 2, "High school graduate",
  some_ps == 1, "Some post-secondary",
  ps_cert == 1, "Post-secondary certificate",
  ba == 1, "Bachelor's degree",
  above_ba == 1, "Above bachelor's",
  default = NA_character_
)]
education_subsets <- split(LFS_data, LFS_data$education_level)
education_gini_results <- data.frame(
  group = character(),
  years_from_start = numeric(),
  gini = numeric(),
  median_income = numeric(),
  stringsAsFactors = FALSE
)

for (education_name in names(education_subsets)) {
  if (is.na(education_name)) next
  
  subset_gini <- calculate_gini_by_year(education_subsets[[education_name]])
  
  if (nrow(subset_gini) > 0) {
    subset_gini_df <- data.frame(
      group = education_name,
      years_from_start = subset_gini$years_from_start,
      gini = subset_gini$gini,
      median_income = subset_gini$median_income,
      stringsAsFactors = FALSE
    )
    education_gini_results <- rbind(education_gini_results, subset_gini_df)
  }
}

# Convert to actual years and create education plot
education_gini_results$year <- 2006 + education_gini_results$years_from_start
filtered_education_gini <- education_gini_results[!is.na(education_gini_results$group), ]
p <- ggplot(filtered_education_gini, aes(x = year, y = gini, color = group)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, alpha = 0.2) +
  labs(title = "Income Inequality Trends by Education Level",
       x = "Year",
       y = "Gini Coefficient",
       color = "Education Level") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_y_continuous(limits = c(0.15, 0.4))
print(p)
rm(p, education_subsets, education_gini_results, filtered_education_gini, subset_gini, subset_gini_df, education_name)
invisible(gc())

# Process gender-education intersectional analysis
LFS_data[, gender_education := paste(named_sex, education_level, sep = " - ")]
gender_education_subsets <- split(LFS_data, LFS_data$gender_education)
gender_education_gini_results <- data.frame(
  group = character(),
  years_from_start = numeric(),
  gini = numeric(),
  median_income = numeric(),
  stringsAsFactors = FALSE
)

for (gender_education_name in names(gender_education_subsets)) {
  if (grepl("NA", gender_education_name)) next
  
  subset_gini <- calculate_gini_by_year(gender_education_subsets[[gender_education_name]])
  
  if (nrow(subset_gini) > 0) {
    subset_gini_df <- data.frame(
      group = gender_education_name,
      years_from_start = subset_gini$years_from_start,
      gini = subset_gini$gini,
      median_income = subset_gini$median_income,
      stringsAsFactors = FALSE
    )
    gender_education_gini_results <- rbind(gender_education_gini_results, subset_gini_df)
  }
  
  # Periodic memory cleanup
  if (nrow(gender_education_gini_results) %% 10 == 0) {
    invisible(gc())
  }
}

# Convert to actual years and create intersectional plots
gender_education_gini_results$year <- 2006 + gender_education_gini_results$years_from_start

# Filter to key education levels for readability
key_education_levels <- c(
  "Male - High school graduate", 
  "Female - High school graduate",
  "Male - Post-secondary certificate", 
  "Female - Post-secondary certificate",
  "Male - Bachelor's degree", 
  "Female - Bachelor's degree",
  "Male - Above bachelor's", 
  "Female - Above bachelor's"
)

filtered_gender_education <- gender_education_gini_results[
  gender_education_gini_results$group %in% key_education_levels, 
]

# Create Gini coefficient plot
p <- ggplot(filtered_gender_education, aes(x = year, y = gini, color = group)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, alpha = 0.2) +
  labs(title = "Income Inequality by Gender and Education Level",
       x = "Year",
       y = "Gini Coefficient",
       color = "Group") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_y_continuous(limits = c(0.15, 0.4))
print(p)
rm(p)

# Create median income plot
p <- ggplot(filtered_gender_education, aes(x = year, y = median_income, color = group)) +
  geom_line(linewidth = 1.2) +
  geom_smooth(method = "loess", se = FALSE, alpha = 0.2) +
  labs(title = "Median Hourly Income by Gender and Education Level",
       x = "Year",
       y = "Median Hourly Income ($)",
       color = "Group") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_y_continuous(labels = scales::dollar)
print(p)

# Clean up all remaining objects
rm(p, gender_education_subsets, gender_education_gini_results, filtered_gender_education, 
   key_education_levels, subset_gini, subset_gini_df, gender_education_name, calculate_gini_by_year)
if(exists("LFS_data")) {
  rm(LFS_data)
}
invisible(invisible(gc()))
```

The intersectional analysis reveals important dynamics in both inequality (Gini coefficient) and median incomes across gender and education groups. While certain education levels exhibit similar patterns of inequality for both genders, the median income graph clearly demonstrates persistent gender wage gaps at every education level, with the gap widening at higher education levels.

# Limitations of the Data

As this paper uses the publicly available LFS data, certain information is excluded -- there are no observations for the Territories, nor is there any access to ethnicity/race variables. The standard scope of the LFS also applies: full-time members of the armed forces, people living on reserves, institutionalized persons, and those under the age of 15 are excluded.

# Conclusions and Policy Implications

[To be completed after analysis]

# References

[To be added]

# AI Disclosures

Through the course of this project, I utilized Gemini and ChatGPT. Unlike ChatGPT, Gemini does not provide an easy way to share the whole chat, so I've included some sample queries and responses and wrote a quick blurb explaining how the tools were used.

Set-up help: <https://g.co/gemini/share/085ca32bc3d4>, <https://g.co/gemini/share/429555247dcd>

Writing assistance: <https://chatgpt.com/share/67eebc76-c988-8010-801d-fbc4827958f2>

Code assistance: <https://g.co/gemini/share/cc0925dc8fb8>, <https://g.co/gemini/share/6866b019b792> <https://g.co/gemini/share/898697efa94a>

My main workflow for code corrections was to boot up a chat with Gemini and ask for help with specific R syntax issues rather than entire workflow planning. This approach allowed me to maintain control over the overall structure while getting assistance with technical implementation details. Or for general questions, for example this question and response <https://g.co/gemini/share/4acf62d26b19> where I ask how I can make it so that when I call gc() it doesn't put the stats in the knitted doc, and it told me I could just replace every gc() call with invisible(gc())
