---
title: "Data Research Project"
author: "Alex McKeown"
date: "2025-03-29"
output: html_document
---

```{r setup, include=FALSE}
#set up, clear workspace, import libraries, then import data from folder
#this is not a fun section to run, try to run it once then don't run again
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)
# List of required packages
required_packages <- c(
  "dplyr",
  "tidyverse",  # Includes ggplot2, readr, etc.
  "plm",        # For panel data regression models
  "caret",      # For ML tools like cross-validation
  "cluster",    # For clustering algorithms
  "modelsummary", # For summarizing regressions
  "ineq",        # For calculating the Gini coefficient
  "data.table"  # For more efficient merging of CSV files
)

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    message(paste("Installing package:", pkg))
    install.packages(pkg, dependencies = TRUE)
  }
  library(pkg, character.only = TRUE)
  message(paste("Loaded package:", pkg))
}

# Set your working directory to the folder with the CSVs
# IMPORTANT: Replace with your actual path
setwd("/Users/alex/Desktop/ECO340-Project")

# List all CSV files in the folder
csv_files <- list.files(pattern = "\\.csv$", full.names = TRUE) # Use full.names = TRUE for fread

# Check if any CSV files were found
if (length(csv_files) == 0) {
  stop("No CSV files found in the specified directory.")
} else {
  message(paste("Found", length(csv_files), "CSV files. Attempting to read and combine..."))

  # Use data.table::fread for faster, memory-efficient reading
  # Use data.table::rbindlist for efficient binding
  # Wrap in tryCatch to handle potential errors during file reading/binding
  list_of_data_tables <- NULL # Initialize to NULL
  raw_data <- tryCatch({
    # Read each CSV file using fread into a list
    list_of_data_tables <<- lapply(csv_files, data.table::fread) # Use <<- to assign to outer scope variable
    # Combine the list of data.tables into one and return it directly
    message("Successfully read and combined CSV files.")
    data.table::rbindlist(list_of_data_tables) # Directly return the combined table
  }, error = function(e) {
    # Print error message if reading/binding fails
    message("Error reading or binding CSV files: ", e$message)
    return(NULL) # Return NULL or an empty data structure if preferred
  })

  # Optional: Clean up the list of data tables to free memory sooner
  # Check if the list was created before trying to remove it
  if (!is.null(list_of_data_tables)) {
    rm(list_of_data_tables)
    message("Intermediate list 'list_of_data_tables' removed.")
  }
  gc() # Suggest garbage collection

  # Check if data loading was successful
  if (!is.null(raw_data)) {
     print("Data loaded successfully. Dimensions:")
     print(dim(raw_data))
     # print(head(raw_data)) # Uncomment to view the first few rows
  } else {
     print("Data loading failed. Please check file paths, permissions, and file integrity.")
  }
}
```

## Introduction

Over the years, Canada's Gini coefficient with respect to income has fluctuated over the years. This analysis endeavours to show if there have been any meaningful trends in inequities within different age and gender groupings of the Canadian population. The first part of this analysis will be concerned with the Gini coeffients for each age group plotted over the past 24 years, using data from the Canadian LFS from March 2000 through to March 2024. The second part of the analysis will be looking at trends in unemployment and labour force participation over this same period of time for each group.

As this data is not like the NLSY in that it follows the same people over the years, this will not be a generational/cohort study, ie the age brackets will remain fixed and over the years the cohorts will move through them. As well, as this analysis is using the publicly available version of the LFS, data is not provided on race or certain other potentially identifying characteristics.

##Data Cleaning & Variable Creation

To make analysis easier, some new variables will be created to make an easier time of things. Variables which give multiple categories in numeric form, for example province codes, will be converted to binary dummy variables. The variables that will be converted in this fashion are: gender, re-coding from 1 & 2 to 0 & 1; union status, breaking the "union member" "union contract" and "neither" into two dummy variables for "union member" and "union contract"; the aformentioned provinces, creating 9 dummy variables with Ontario being the reference province; education, creating two sets of dummy variables one being a culmulative set (ie each person only has one dummy variable that's in the on position) and the other being individual (ie someone with a college education has college turned on as well as every switch before); age, recoded to be an approximate age using both AGE12 and AGE6 data to maximize precision; LFSSTAT, recoded from a 1, 2, 3, 4 to a "in labour force", "is employed", and "absent from work" dummies; immigrant similarly into two dummies, one for recent immigrant one for established immigrant; year and month data into months from start of data-set (I changed directions so didn't really use this).

```{r}
#For simplicity we will break down the variable LFSSTAT into an Employed binary variable and an in labour force variable. Let's label these is_employed and in_lf -- using these we can more easily down the line calculated changing unemployment rates and labour force participation rates.
#Gender in the LFS datafile is expressed as either a 1 for men and 2 for women, stored as GENDER. We will create a new variable, woman, which will be a binary variable -- 0 for men, 1 for women.
#as opposed to using province codes, I have created new binary variables for the 9 provinces which aren't Ontario. Because I am biased, I will be using Ontario as the default.
#similar to the step for dis-aggregation of the LFSSTAT variable, will decompose the union variable, which gives a 1 for being a union member, a 2 for being covered by union contract, and a 3 for being not covered (an 8 for not being applicable, but dwai bc we won't run a regression on them), into a binary for being a union member (ie 1 only) and a binary for being covered by union contract (1 and 2). we will call these union_member and union_contract (outstanding creativity there w the names)
#same process now for IMMIG, immigration status. going to create variables newer_immigrant for immigrants here for 10 or less years, older_immigrant for those who have been here for more than 10 years.
#education will be broken into culmulative dummies, NOTE TO SELF NEED TO REMEMBER THERE IS SIG COLLINEARITY MUST ACCOUNT FOR, as well as a set of exclusive categories for regression purposes :)

raw_data <- raw_data %>%
  mutate(
    is_employed = ifelse(LFSSTAT == 1, 1, 0),
    in_lf = ifelse(LFSSTAT %in% c(1, 2), 1, 0),
    woman = ifelse(SEX == 2, 1, 0),
    prov_nl = ifelse(PROV == 10, 1, 0),
    prov_pe = ifelse(PROV == 11, 1, 0),
    prov_ns = ifelse(PROV == 12, 1, 0),
    prov_nb = ifelse(PROV == 13, 1, 0),
    prov_qc = ifelse(PROV == 24, 1, 0),
    prov_mb = ifelse(PROV == 46, 1, 0),
    prov_sk = ifelse(PROV == 47, 1, 0),
    prov_ab = ifelse(PROV == 48, 1, 0),
    prov_bc = ifelse(PROV == 59, 1, 0),
    union_member = ifelse(UNION == 1, 1, 0),
    union_contract = ifelse(UNION %in% c(1, 2), 1, 0),
    newer_immigrant = ifelse(IMMIG == 1, 1, 0),
    older_immigrant = ifelse(IMMIG == 2, 1, 0),
    #Culmulative Ed
    some_hs_plus = ifelse(EDUC >= 1, 1, 0),
    hs_grad_plus = ifelse(EDUC >= 2, 1, 0),
    some_ps_plus = ifelse(EDUC >= 3, 1, 0),
    ps_cert_plus = ifelse(EDUC >= 4, 1, 0),
    ba_plus = ifelse(EDUC >= 5, 1, 0),
    above_ba_plus = ifelse(EDUC == 6, 1, 0),
    #Exclusive Ed
    some_hs = ifelse(EDUC == 1, 1, 0),
    hs_grad = ifelse(EDUC == 2, 1, 0),
    some_ps = ifelse(EDUC == 3, 1, 0),
    ps_cert = ifelse(EDUC == 4, 1, 0),
    ba = ifelse(EDUC == 5, 1, 0),
    above_ba = ifelse(EDUC == 6, 1, 0)
  )

#going to create a variable which is a mid-point estimate for age, combining variables AGE_12 and AGE_6, this new variable will be called age_rough.

raw_data <- raw_data %>%
  mutate(age_rough = case_when(
    AGE_6 == 1 ~ 15.5,      # 15–16 (from AGE_6)
    AGE_6 == 2 ~ 18,        # 17–19 (from AGE_6)
    AGE_12 == 2 ~ 22,       # 20–24
    AGE_12 == 3 ~ 27,       # 25–29
    AGE_12 == 4 ~ 32,       # 30–34
    AGE_12 == 5 ~ 37,       # 35–39
    AGE_12 == 6 ~ 42,       # 40–44
    AGE_12 == 7 ~ 47,       # 45–49
    AGE_12 == 8 ~ 52,       # 50–54
    AGE_12 == 9 ~ 57,       # 55–59
    AGE_12 == 10 ~ 62,      # 60–64
    AGE_12 == 11 ~ 67,      # 65–69
    AGE_12 == 12 ~ 75       # 70+
  ))

# Create a variable for months from start of data (starting from January 2014)
raw_data <- raw_data %>%
  mutate(months_from_start = (SURVYEAR - 2006) * 12 + SURVMNTH)

```

## Creating our sets of Gini Coefficients

now we want to create time series of Gini cofficients for each of our "groups" of Canadians. First let's just quickly demonstrate the process by calculating the Gini coefficient for our whole dataset over the time series, using values for HRLYEARN. 

```{r}
# Ensure necessary packages are loaded (dplyr, tidyr, ineq, ggplot2)
# These should be handled by the r_package_check_install artifact if run first
# library(dplyr)
# library(tidyr)
# library(ineq)
# library(ggplot2)
# library(lubridate) # Potentially useful for date calculations, though base R used here

# --- Create Age Group Labelling ---
# Assuming 'raw_data' is your data frame with AGE_6 and AGE_12 columns
raw_data <- raw_data %>%
  mutate(
    age_group = case_when(
      # Ages 15-19 based on AGE_6
      AGE_6 %in% c(1, 2) ~ "15-19",
      # Ages 20-34 based on AGE_12
      AGE_12 %in% c(2, 3, 4) ~ "20-34",
      # Ages 35-54 based on AGE_12
      AGE_12 %in% c(5, 6, 7, 8) ~ "35-54",
      # Ages 55+ based on AGE_12
      AGE_12 %in% c(9, 10, 11, 12) ~ "55+",
      # Optional: Handle cases that don't fit any category (e.g., NA or other values)
      TRUE ~ NA_character_ # Assign NA if none of the above conditions are met
    ),
    # Optional: Ensure the new column is a factor if preferred for analysis/plotting
    # age_group = factor(age_group, levels = c("15-19", "20-34", "35-54", "55+"))
  )

# --- Calculate Gini Coefficient Time Series by Age Group ---
# Assuming 'raw_data' also contains 'months_from_start' and 'HRLYEARN'

# Check if the required columns exist (add checks if not already present)
if (!all(c("months_from_start", "HRLYEARN", "age_group") %in% names(raw_data))) {
  stop("Error: One or more required columns (months_from_start, HRLYEARN, age_group) not found in raw_data.")
}
# Ensure 'HRLYEARN' is numeric (add conversion if needed)
if (!is.numeric(raw_data$HRLYEARN)) {
   # Add robust numeric conversion here if not already done
   warning("HRLYEARN column is not numeric. Ensure it's converted before this step.")
}


gini_timeseries <- raw_data %>%
  # Filter out rows where age_group or HRLYEARN is NA, as they can't be used
  filter(!is.na(age_group), !is.na(HRLYEARN)) %>%
  # Group data by the time period identifier AND age group
  group_by(months_from_start, age_group) %>%
  # Calculate the Gini coefficient for each subgroup
  summarize(
    # Use the Gini function from the 'ineq' package
    gini_value = ineq::Gini(HRLYEARN, na.rm = TRUE), # na.rm = TRUE is still good practice
    # Optional: Add a count of non-NA observations used for each calculation
    n_observations = n(), # Use n() since we filtered NAs earlier
    .groups = 'drop' # Drop grouping structure after summarize
  ) %>%
  # Reshape data from long to wide format
  tidyr::pivot_wider(
    id_cols = months_from_start, # Column(s) to keep as identifiers
    names_from = age_group,      # Column whose values become new column names
    values_from = c(gini_value, n_observations), # Column(s) whose values fill the new columns
    names_sep = "_",             # Separator between value name and age group name
    # Corrected prefix application: Apply prefix within names_glue or rename later
    # Using names_glue for more control:
    names_glue = "{age_group}_{.value}" # Creates names like 15-19_gini_value, 15-19_n_observations
  ) %>%
  # Arrange by time period for clarity
  arrange(months_from_start)

# --- Plot Gini Coefficients Over Time by Age Group ---

# Check if gini_timeseries was created successfully and has data
if (exists("gini_timeseries") && nrow(gini_timeseries) > 0) {

  # Reshape the data back to long format specifically for plotting Gini values
  # AND add a Date column based on months_from_start (assuming 1 = Jan 2014)
  gini_timeseries_long_for_plot <- gini_timeseries %>%
    # Select the time column and columns ending with '_gini_value'
    select(months_from_start, ends_with("_gini_value")) %>%
    pivot_longer(
      cols = ends_with("_gini_value"), # Select columns to lengthen
      names_to = "age_group",          # New column for the original column names (e.g., "15-19_gini_value")
      names_pattern = "(.*)_gini_value", # Extract the age group part before "_gini_value"
      values_to = "gini_coefficient"   # New column for the Gini values
    ) %>%
    # Add a Date column
    mutate(
      # Calculate year and month based on months_from_start
      year = 2014 + floor((months_from_start - 1) / 12),
      month = ((months_from_start - 1) %% 12) + 1,
      # Create a Date object for the first day of the calculated month/year
      plot_date = as.Date(paste(year, month, 1, sep = "-"))
    )

  # Create the plot using ggplot2, mapping plot_date to x-axis
  gini_plot <- ggplot(gini_timeseries_long_for_plot,
                      aes(x = plot_date, y = gini_coefficient, color = age_group)) +
    geom_line(linewidth = 1) + # Thicker line for visibility
    geom_point(size = 1.5) +    # Add points
    # Use scale_x_date for better control over date axis formatting
    scale_x_date(
        date_breaks = "1 year", # Major breaks every year
        date_minor_breaks = "3 months", # Minor breaks every 3 months
        date_labels = "%Y" # Display only the year for major breaks
        # Alternative label formats: "%b %Y" (e.g., Jan 2014), "%Y-%m" (e.g., 2014-01)
    ) +
    labs(
      title = "Gini Coefficient of Hourly Earnings Over Time by Age Group",
      x = "Date", # Updated x-axis label
      y = "Gini Coefficient",
      color = "Age Group" # Legend title
    ) +
    theme_minimal(base_size = 12) + # Use a clean theme with slightly larger base font
    theme(legend.position = "bottom") # Move legend below plot

  # Print the plot to display it
  print(gini_plot)

} else {
  message("Skipping plot: 'gini_timeseries' data frame not found or is empty.")
}



```

## Findings from simple regressions

Here we see some evidence that women make less, on average, then men do in the Canadian labour force. The aim of this paper is not to investigate whether and how there is gender discrimination, but rather to determine whether the gender wage gap varies with the business cycle -- during recessionary periods do we see women get laid off first? Let's build a separate data frame for our "macro" variables: % change in Real GDP, % change in Real GDP per Capita, and just for fun lets also throw in the inflation rate, a measure of inequality (gini coefficient), and overall labour participation and unemployment rates to compare our LFS data to. We can then map on these values into our original dataset -- for example, gdp growth data is reported quarterly, we can mutate a new variable in such that if year == 20xx and month == c(a, b, c) = growth value for that quarter.

``` {r}

```

## Limitations of the Data

As this paper is using the publicly available LFS data, certain information is excluded -- thanks to this there are no observations for the Territories, nor is there any access to ethnicity/race variables. Finally, the standard scope of the LFS applies: full time members of the armed forces, people living on reserves, institutionalized persons, and those under the age of 15 are excluded.

## ChatGPT Disclosure:

Through the course of this project, I utilized Gemini. Below are chat links, organized by theme/part of the project the chat is relevant:

set-up: <https://g.co/gemini/share/085ca32bc3d4>
This was crucial, only noticed that I was doing a very inefficient way to merge the csv files when I swapped from working on my desktop with 64Gb of RAM to my MacBook with 16Gb lol. I also used it to make a little loop function for loading the packages, just to make it easier to add new packages I want to use as I go through the project.

general writing: <https://chatgpt.com/share/67eebc76-c988-8010-801d-fbc4827958f2>

general code and data-set questions: <https://chatgpt.com/share/67eebc87-d4a0-8010-9c20-69841f956c4d>
