---
title: "Income Inequality and Labour Market Outcomes in Canada, 2000-2024"
author: "Alex McKeown"
date: "2025-03-29"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    fig_caption: true
    number_sections: true
    theme: cosmo
    css: styles.css
    highlight: tango
    code_folding: hide
    self_contained: true
    df_print: paged
    mathjax: default
---

```{css, echo=FALSE}
body {
  font-family: "Times New Roman", Times, serif;
  line-height: 2;
  font-size: 12pt;
}

h1, h2, h3, h4, h5, h6 {
  font-family: "Times New Roman", Times, serif;
  font-weight: bold;
}

.main-container {
  max-width: 8.5in;
  margin: auto;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,       # Changed from FALSE to TRUE to allow code folding
  warning = FALSE,    
  message = FALSE,    
  fig.width = 10,     
  fig.height = 6,     
  fig.align = "center", 
  dpi = 300           
)

# Make sure the required packages for code folding are loaded
if (!requireNamespace("rmarkdown", quietly = TRUE)) {
  install.packages("rmarkdown")
}
if (!requireNamespace("knitr", quietly = TRUE)) {
  install.packages("knitr")
}
```

```{r packages-and-data, include=FALSE}
# Setup - Clear workspace and manage packages
# Define required packages, install if needed, load all packages, set working directory,
# load and combine data from CSV files matching the pattern "^pub(01|06)\\d{2}\\.csv$"
rm(list = ls())

required_packages <- c(
  "data.table", "dplyr", "tidyverse", "plm",
  "caret", "cluster", "modelsummary", "ineq", "ggplot2", "broom", 
  "rmarkdown"
)

packages_loaded <- sapply(required_packages, function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    message(paste("Installing package:", pkg))
    tryCatch({
        install.packages(pkg, dependencies = TRUE)
        library(pkg, character.only = TRUE)
        return(TRUE)
      }, error = function(e) {
        message(paste("Failed to install package:", pkg, "Error:", e$message))
        return(FALSE)
      })
  } else {
    library(pkg, character.only = TRUE)
    return(TRUE)
  }
})

if (!packages_loaded["data.table"]) {
  stop("Critical package 'data.table' could not be loaded. Aborting.")
}

csv_folder_path <- "/Users/alex/Desktop/ECO340-Project" # USER: MODIFY THIS PATH
tryCatch({
    setwd(csv_folder_path)
  }, warning = function(w){
    stop("Warning setting working directory: ", w$message, ". Check path: ", csv_folder_path)
  }, error = function(e) {
    stop("Error setting working directory: ", e$message, ". Check path: ", csv_folder_path)
  }
)

# Find and load CSV files matching the pattern
all_csv_files <- list.files(path = getwd(), pattern = "\\.csv$", full.names = TRUE, ignore.case = TRUE)
filename_pattern <- "^pub(01|06)\\d{2}\\.csv$"
all_basenames <- basename(all_csv_files)
matching_indices <- grepl(filename_pattern, all_basenames, ignore.case = TRUE)
csv_files <- all_csv_files[matching_indices]

if (length(csv_files) == 0) {
  stop("No CSV files matching the pattern '", filename_pattern, "' were found in ", getwd())
} else {
  LFS_data <- NULL
  for (i in seq_along(csv_files)) {
    file_path <- csv_files[i]

    current_data <- tryCatch({
        data.table::fread(file_path, showProgress = FALSE)
      }, error = function(e) {
        return(NULL)
      })

    if (!is.null(current_data) && nrow(current_data) > 0) {
      if (is.null(LFS_data)) {
        LFS_data <- current_data
      } else {
        LFS_data <- data.table::rbindlist(list(LFS_data, current_data), use.names = TRUE, fill = TRUE)
      }
    }
    gc()
  }
  gc()
}
```

# Introduction

Over the years, Canada's Gini coefficient with respect to income has fluctuated. This analysis endeavours to show if there have been any meaningful trends in inequities within different age and gender groupings of the Canadian population. The first part of this analysis will be concerned with the Gini coeffients for each age group plotted over the past 24 years, using data from the Canadian Labour Force Survey (LFS) from March 2000 through to March 2024. The second part of the analysis will be looking at trends in unemployment and labour force participation over this same period of time for each group.

As this data is not like the NLSY in that it follows the same people over the years, this will not be a generational/cohort study, i.e., the age brackets will remain fixed and over the years the cohorts will move through them. As well, as this analysis is using the publicly available version of the LFS, data is not provided on race or certain other potentially identifying characteristics.

# Data Cleaning and Variable Creation

```{r data-cleaning, include=FALSE}
# Data Cleaning & Variable Creation
# Convert variables to appropriate formats and create binary indicators for analysis

setDT(LFS_data)

#Log Hourly Earnings
LFS_data[, ln_income := log(HRLYEARN)]

# Employment and Labour Force Status
LFS_data[, is_employed := fifelse(LFSSTAT == 1, 1, 0)]
LFS_data[, in_lf := fifelse(LFSSTAT %in% c(1, 2), 1, 0)]

# Sex
LFS_data[, woman := fifelse(SEX == 2 | GENDER == 2, 1L, 0L, na = 0L)]

# Age
LFS_data[, rough_age_dock15 := fcase(
    AGE_12 == 1,  2, AGE_12 == 2,  7, AGE_12 == 3,  12, AGE_12 == 4,  17,
    AGE_12 == 5,  22, AGE_12 == 6,  27, AGE_12 == 7,  32, AGE_12 == 8,  37,
    AGE_12 == 9,  42, AGE_12 == 10, 47, AGE_12 == 11, 52, AGE_12 == 12, 57,
    default = NA_real_
)]

# Province indicators
LFS_data[, prov_nl := fifelse(PROV == 10, 1, 0)]
LFS_data[, prov_pe := fifelse(PROV == 11, 1, 0)]
LFS_data[, prov_ns := fifelse(PROV == 12, 1, 0)]
LFS_data[, prov_nb := fifelse(PROV == 13, 1, 0)]
LFS_data[, prov_qc := fifelse(PROV == 24, 1, 0)]
LFS_data[, prov_mb := fifelse(PROV == 46, 1, 0)]
LFS_data[, prov_sk := fifelse(PROV == 47, 1, 0)]
LFS_data[, prov_ab := fifelse(PROV == 48, 1, 0)]
LFS_data[, prov_bc := fifelse(PROV == 59, 1, 0)]
province_dummies <- c("prov_nl", "prov_pe", "prov_ns", "prov_nb", "prov_qc", "prov_mb", "prov_sk", "prov_ab", "prov_bc")

# Union status
LFS_data[, union_member := fifelse(UNION == 1, 1, 0)]
LFS_data[, union_contract := fifelse(UNION %in% c(1, 2), 1, 0)]

# Immigration status
LFS_data[, newer_immigrant := fifelse(IMMIG == 1, 1, 0)]
LFS_data[, older_immigrant := fifelse(IMMIG == 2, 1, 0)]

# Education dummies

LFS_data[, zero_to_eight := fifelse(EDUC == 0, 1, 0)]
LFS_data[, some_hs := fifelse(EDUC == 1, 1, 0)]
LFS_data[, some_ps := fifelse(EDUC == 3, 1, 0)]
LFS_data[, ps_cert := fifelse(EDUC == 4, 1, 0)]
LFS_data[, ba := fifelse(EDUC == 5, 1, 0)]
LFS_data[, above_ba := fifelse(EDUC == 6, 1, 0)]
education_dummies <- c("zero_to_eight", "some_hs", "some_ps", "ps_cert", "ba", "above_ba")

# NAICS industry dummies
LFS_data[, naics_ForestryLogSupport := fifelse(NAICS_21 == 2, 1, 0)]
LFS_data[, naics_FishingHuntTrap := fifelse(NAICS_21 == 3, 1, 0)]
LFS_data[, naics_MiningOilGasExtract := fifelse(NAICS_21 == 4, 1, 0)]
LFS_data[, naics_Utilities := fifelse(NAICS_21 == 5, 1, 0)]
LFS_data[, naics_Construction := fifelse(NAICS_21 == 6, 1, 0)]
LFS_data[, naics_MfgDurable := fifelse(NAICS_21 == 7, 1, 0)]
LFS_data[, naics_MfgNonDurable := fifelse(NAICS_21 == 8, 1, 0)]
LFS_data[, naics_Wholesale := fifelse(NAICS_21 == 9, 1, 0)]
LFS_data[, naics_Retail := fifelse(NAICS_21 == 10, 1, 0)]
LFS_data[, naics_TransportWarehousing := fifelse(NAICS_21 == 11, 1, 0)]
LFS_data[, naics_FinanceInsurance := fifelse(NAICS_21 == 12, 1, 0)]
LFS_data[, naics_RealEstateRentalLeasing := fifelse(NAICS_21 == 13, 1, 0)]
LFS_data[, naics_ProfSciTech := fifelse(NAICS_21 == 14, 1, 0)]
LFS_data[, naics_BusinessBuildingSupport := fifelse(NAICS_21 == 15, 1, 0)]
LFS_data[, naics_Educational := fifelse(NAICS_21 == 16, 1, 0)]
LFS_data[, naics_HealthSocialAssist := fifelse(NAICS_21 == 17, 1, 0)]
LFS_data[, naics_InfoCultRec := fifelse(NAICS_21 == 18, 1, 0)]
LFS_data[, naics_AccomFood := fifelse(NAICS_21 == 19, 1, 0)]
LFS_data[, naics_OtherServices := fifelse(NAICS_21 == 20, 1, 0)]
LFS_data[, naics_PublicAdmin := fifelse(NAICS_21 == 21, 1, 0)]
naics_dummies <- c(
  "naics_ForestryLogSupport", "naics_FishingHuntTrap", "naics_MiningOilGasExtract", "naics_Utilities",
  "naics_Construction", "naics_MfgDurable", "naics_MfgNonDurable", "naics_Wholesale",
  "naics_Retail", "naics_TransportWarehousing", "naics_FinanceInsurance", "naics_RealEstateRentalLeasing",
  "naics_ProfSciTech", "naics_BusinessBuildingSupport", "naics_Educational", "naics_HealthSocialAssist",
  "naics_InfoCultRec", "naics_AccomFood", "naics_OtherServices", "naics_PublicAdmin"
)

# Time variables
LFS_data[, years_from_start := (SURVYEAR - 2006) + ifelse(SURVMNTH == 6, 0.5, 0)]
yfs_dummy_names <- character(0)
unique_vals <- sort(unique(na.omit(LFS_data$years_from_start)))

if (length(unique_vals) > 0) {
  for (val in unique_vals) {
    new_col_name <- paste0("yfs_", gsub("[^a-zA-Z0-9_]", "_", as.character(val)))
    yfs_dummy_names <- c(yfs_dummy_names, new_col_name)
    LFS_data[, (new_col_name) := fifelse(years_from_start == val, 1L, 0L, na = 0L)]
  }
}

# Categorical variables for grouping
LFS_data[, province_name := fcase(
    PROV == 10, "Newfoundland and Labrador", PROV == 11, "Prince Edward Island",
    PROV == 12, "Nova Scotia", PROV == 13, "New Brunswick",
    PROV == 24, "Quebec", PROV == 35, "Ontario",
    PROV == 46, "Manitoba", PROV == 47, "Saskatchewan",
    PROV == 48, "Alberta", PROV == 59, "British Columbia"
)]

LFS_data[, named_sex := fcase(woman == 0, "Male", woman == 1, "Female")]

LFS_data[, age_group := fcase(
    AGE_12 == 1, "15_19",
    AGE_12 %in% c(2, 3, 4), "20_34",
    AGE_12 %in% c(5, 6, 7, 8), "35_54",
    AGE_12 %in% c(9, 10, 11, 12), "55+",
    default = NA_character_
)]

gc()
```

To make analysis easier, I created several derived variables from the raw LFS data. Variables with multiple categories in numeric form, such as province codes, were converted to binary dummy variables. The key transformations include:

- Gender: re-coded from 1 & 2 to 0 & 1 for regression analysis
- Union status: created separate indicators for union membership and being covered by a union contract
- Provinces: created 9 dummy variables with Ontario as the reference province
- Education: created dummy variables, one for each level of education
- Age: approximated using AGE12 variable, creating a running variable and creating four age brackets for dividing up sample later on (15-19, 20-34, 35-54, and 55+)
- Labour force status: converted LFSSTAT into binary indicators for employment and labour force participation
- Immigration status: created indicators for recent immigrants (≤10 years) and established immigrants
- Income: created a logarithmic version of hourly earnings for easier interpretation in regressions

These transformations allow for more efficient analysis and interpretation of results in subsequent sections.

# Wage Determinants: Regression Analysis

This section examines the impact of key variables and controls on hourly earnings. As this model is estimated by applying it to multiple waves of LFS data, year dummy variables are included to account for time fixed-effects and year-specific events. The analysis is restricted to employed Canadians, with variations in unemployment and labour force participation covered later.

The purpose of this section is not to derive any profound insights into the determinants of wages in Canada, rather the goal is to create a starting point for how we can divide up our sample to compare the trends of Gini coefficents over time.

The dependent variable in these regressions is the natural logarithm of hourly earnings, for easier interpretation of coefficients. The "base" or intercept observation is a male high school graduate from Ontario who is not in a union, nor is covered by a union contract, nor is an immigrant, and works in the Agriculture sector.

```{r wage-regressions}
# Store the models for comparison
models <- list()

# Model 1: Basic with gender and year effects only
dummy_formula_part <- paste(yfs_dummy_names, collapse = " + ")
regression_formula <- as.formula(paste("ln_income ~ woman +", dummy_formula_part))
models[[1]] <- lm(regression_formula, data = LFS_data)

# Model 2: Adding province effects
dummy_formula_part1 <- paste(yfs_dummy_names, collapse = " + ")
dummy_formula_part2 <- paste(province_dummies, collapse = " + ")
regression_formula <- as.formula(paste("ln_income ~ woman + ", dummy_formula_part1, " + ", dummy_formula_part2))
models[[2]] <- lm(regression_formula, data = LFS_data)

# Model 3: Adding education effects
dummy_formula_part1 <- paste(yfs_dummy_names, collapse = " + ")
dummy_formula_part2 <- paste(province_dummies, collapse = " + ")
dummy_formula_part3 <- paste(education_dummies, collapse = " + ")
regression_formula <- as.formula(paste("ln_income ~ woman + ", dummy_formula_part1, " + ", dummy_formula_part2, " + ", dummy_formula_part3))
models[[3]] <- lm(regression_formula, data = LFS_data)

# Model 4: Adding age polynomial
dummy_formula_part1 <- paste(yfs_dummy_names, collapse = " + ")
dummy_formula_part2 <- paste(province_dummies, collapse = " + ")
dummy_formula_part3 <- paste(education_dummies, collapse = " + ")
regression_formula <- as.formula(paste("ln_income ~ woman + poly(rough_age_dock15, 2, raw=TRUE) + ",
                                     dummy_formula_part1, " + ", dummy_formula_part2, " + ", dummy_formula_part3))
models[[4]] <- lm(regression_formula, data = LFS_data)

# Model 5: Full model with union, immigration, and industry sectors
dummy_formula_part1 <- paste(yfs_dummy_names, collapse = " + ")
dummy_formula_part2 <- paste(province_dummies, collapse = " + ")
dummy_formula_part3 <- paste(education_dummies, collapse = " + ")
dummy_formula_part4 <- paste(naics_dummies, collapse = " + ")
regression_formula <- as.formula(paste("ln_income ~ woman + poly(rough_age_dock15, 2, raw=TRUE) + newer_immigrant + older_immigrant + union_contract + ",
                                     dummy_formula_part1, " + ", dummy_formula_part2, " + ", dummy_formula_part3, " + ", dummy_formula_part4))
models[[5]] <- lm(regression_formula, data = LFS_data)

# Define model names for the table
model_names <- c("Gender Only", "Add Provinces", "Add Education", "Add Age", "Full Model")
names(models) <- model_names

# Create a custom coefficient map for the table
coef_map <- c(
  "woman" = "Female",
  "poly(rough_age_dock15, 2, raw = TRUE)1" = "Age",
  "poly(rough_age_dock15, 2, raw = TRUE)2" = "Age²",
  "newer_immigrant" = "Recent Immigrant (≤10yrs)",
  "older_immigrant" = "Established Immigrant (>10yrs)",
  "union_contract" = "Union Coverage",
  
  # Key education variables
  "zero_to_eight" = "Education: 0-8 years",
  "ba" = "Education: Bachelor's degree",
  "above_ba" = "Education: Above bachelor's",
  
  # Selected provinces
  "prov_qc" = "Quebec",
  "prov_ab" = "Alberta",
  "prov_bc" = "British Columbia",
  
  # Selected industry sectors
  "naics_MiningOilGasExtract" = "Industry: Mining & Oil Extraction",
  "naics_FinanceInsurance" = "Industry: Finance & Insurance",
  "naics_HealthSocialAssist" = "Industry: Healthcare & Social Assist."
)

# Generate regression table
modelsummary(models,
             title = "Log Wage Regressions with Different Specifications",
             stars = TRUE,
             coef_map = coef_map,
             gof_map = c("nobs", "r.squared", "adj.r.squared"),
             notes = "Year fixed effects included in all models. Ontario is the reference province. Full range of dummy variables not shown for ease of viewing",
             output = "html")
```

## Interpretation of Results

The regression results above show how different factors influence hourly wages in Canada. Key observations include:

1. **Gender Wage Gap**: Across all specifications, women earn less than men, though the gap changes a bit depending on what controls are added. As this gap is persistent, for our later analysis of Gini coefficients between groups, it seems wise to create different groups for women and men.

2. **Provincial Variation**: Substantial differences exist across provinces, with Alberta showing a wage premium and Quebec showing somewhat lower wages compared to Ontario (the reference province). While the Federal government through income taxes and other means of taxation has a large role to play in combating inequality, it seems pertinent to divide by provinces as well, as there are differences in provincial services, cost of living, sectors, etc.

3. **Education Premium**: Higher education levels are associated with substantially higher wages, with the greatest returns to Bachelor's and advanced degrees. This seems like a source of inequity between workers which we do want to capture.

4. **Age Effects**: Wages follow a concave pattern with age, increasing at a decreasing rate and eventually plateauing. If inequalities are caused by life cycles, young workers make less than older workers for example, that's not too problematic to most, young or old, because the young workers will one day be those older workers. Thus in our analysis of Gini coefficients, we will be dividing into age groups. 

5. **Immigration Status**: Recent immigrants face a wage penalty that persists even after controlling for education and industry. This may be due a variety of factors, we will examine this later on. 

6. **Union Coverage**: Union contracts are associated with higher wages, reflecting the collective bargaining premium. This does not seem like an important distinction to make in analysis of Gini coefficients -- if say, union membership is on the decline, but unions are still powerful and give the fewer workers in them a much greater wage, we would want to see that in data for Gini coefficients of say 25-34 year olds. 

7. **Industry Effects**: Significant wage differences exist across industries, with Mining & Oil Extraction and Finance & Insurance offering premiums compared to the Agriculture sector (the reference industry). For the same reason given as with education premiums, this is something which should be included when looking at Gini coefficients for certain segments.

# Income Inequality Analysis

This section examines trends in income inequality across different demographic groups in Canada. Using the Gini coefficient as a measure of inequality, I analyze how wage dispersion has evolved over time for each age, gender, and provincial subgroup.

```{r group-creation, include=FALSE}
gc()
rm(models) # Memory management

# Create nested subsets for analysis
subsetted_data <- split(LFS_data, LFS_data$province_name)
subsetted_data <- lapply(subsetted_data, function(subset_prov) {
  split(subset_prov, subset_prov$named_sex)
})
subsetted_data <- lapply(subsetted_data, function(subset_prov) {
  lapply(subset_prov, function(subset_prov_sex){
    split(subset_prov_sex, subset_prov_sex$age_group)
  })
})
```

```{r gini-calculation}
# Function to calculate Gini coefficient for each subset by year
calculate_gini_by_year <- function(data_subset) {
  if (is.null(data_subset) || nrow(data_subset) < 10) {
    return(data.frame(years_from_start = numeric(0), gini = numeric(0)))
  }
  
  result <- data_subset[!is.na(HRLYEARN), .(
    gini = ineq::ineq(HRLYEARN, type="Gini"),
    count = .N
  ), by = years_from_start]
  
  result <- result[count >= 30]
  return(result[, .(years_from_start, gini)])
}

# Initialize results dataframe
gini_results <- data.frame(
  province = character(),
  sex = character(),
  age_group = character(),
  years_from_start = numeric(),
  gini = numeric()
)

# Process each subset and calculate Gini
for (province_name in names(subsetted_data)) {
  for (sex_name in names(subsetted_data[[province_name]])) {
    for (age_name in names(subsetted_data[[province_name]][[sex_name]])) {
      subset_gini <- calculate_gini_by_year(
        subsetted_data[[province_name]][[sex_name]][[age_name]]
      )
      
      if (nrow(subset_gini) > 0) {
        subset_gini_df <- data.frame(
          province = province_name,
          sex = sex_name,
          age_group = age_name,
          years_from_start = subset_gini$years_from_start,
          gini = subset_gini$gini
        )
        gini_results <- rbind(gini_results, subset_gini_df)
      }
      
      # Periodic memory cleanup
      if (nrow(gini_results) %% 100 == 0) gc()
    }
  }
}

# Convert to actual years for visualization
gini_results$year <- 2006 + gini_results$years_from_start

# Save results to file
write.csv(gini_results, "gini_coefficients_by_group.csv", row.names = FALSE)
```

## National Trends in Income Inequality

The following chart shows how income inequality has changed over time for different gender and age groups across Canada:

```{r national-gini}
# Create national time series
ggplot(gini_results, aes(x = year, y = gini, color = sex, linetype = age_group)) +
  stat_summary(fun = mean, geom = "line") +
  labs(title = "Income Inequality (Gini) Trends by Gender and Age",
       subtitle = "All Provinces Combined",
       x = "Year",
       y = "Gini Coefficient",
       color = "Gender",
       linetype = "Age Group") +
  theme_minimal() +
  scale_y_continuous(limits = c(0.2, 0.3))
```

## Provincial Patterns in Inequality

Next, we examine provincial variations in income inequality:

```{r provincial-gini}
# Create time series by province
ggplot(gini_results, aes(x = year, y = gini, color = sex, linetype = age_group)) +
  stat_summary(fun = mean, geom = "line") +
  facet_wrap(~ province) +
  labs(title = "Income Inequality Trends by Province, Gender, and Age",
       x = "Year",
       y = "Gini Coefficient",
       color = "Gender",
       linetype = "Age Group") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(limits = c(0.2, 0.5))
```

## Age and Gender-Specific Inequality Trends

```{r age-group-gini}
# Create time series by age group (excluding youngest group)
ggplot(gini_results[gini_results$age_group != "15_19", ], aes(x = year, y = gini, color = province)) +
  geom_line() +
  facet_grid(sex ~ age_group) +
  labs(title = "Income Inequality Trends by Age Group and Gender",
       subtitle = "Excludes 15-19 age group with high volatility",
       x = "Year",
       y = "Gini Coefficient",
       color = "Province") +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Focus on Working-Age Population

The working-age population (20-54) shows particularly interesting patterns:

```{r working-age-gini}
# Focus on working-age population
working_age <- gini_results[gini_results$age_group %in% c("20_34", "35_54"), ]

ggplot(working_age, aes(x = year, y = gini, color = sex)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.2) +
  labs(title = "Income Inequality Trends in Working-Age Population (20-54)",
       x = "Year",
       y = "Gini Coefficient",
       color = "Gender") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")
```

# Limitations of the Data

As this paper uses the publicly available LFS data, certain information is excluded -- there are no observations for the Territories, nor is there any access to ethnicity/race variables. The standard scope of the LFS also applies: full-time members of the armed forces, people living on reserves, institutionalized persons, and those under the age of 15 are excluded.

# Conclusions and Policy Implications

[To be completed after analysis]

# References

[To be added]

# AI Disclosures

Through the course of this project, I utilized Gemini and ChatGPT. Unlike ChatGPT, Gemini does not provide an easy way to share the whole chat, so I've included some sample queries and responses and wrote a quick blurb explaining how the tools were used.

Set-up help: <https://g.co/gemini/share/085ca32bc3d4>, <https://g.co/gemini/share/429555247dcd>

Writing assistance: <https://chatgpt.com/share/67eebc76-c988-8010-801d-fbc4827958f2>

Code assistance: <https://g.co/gemini/share/cc0925dc8fb8>, <https://g.co/gemini/share/6866b019b792> <https://g.co/gemini/share/898697efa94a>

My main workflow for code corrections was to boot up a chat with Gemini and ask for help with specific R syntax issues rather than entire workflow planning. This approach allowed me to maintain control over the overall structure while getting assistance with technical implementation details.