---
title: "Data Research Project"
author: "Alex McKeown"
date: "2025-03-29"
output: html_document
---

```{r echo=FALSE}
# Setup - Clear workspace and manage packages
# Define required packages, install if needed, load all packages, set working directory,
# load and combine data from CSV files matching the pattern "^pub(01|06)\\d{2}\\.csv$"
rm(list = ls())

required_packages <- c(
  "data.table", "dplyr", "tidyverse", "plm",
  "caret", "cluster", "modelsummary", "ineq", "ggplot2", "broom"
)

message("--- Loading Packages ---")
packages_loaded <- sapply(required_packages, function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    message(paste("Installing package:", pkg))
    tryCatch({
        install.packages(pkg, dependencies = TRUE)
        library(pkg, character.only = TRUE)
        message(paste("Successfully installed and loaded:", pkg))
        return(TRUE)
      }, error = function(e) {
        message(paste("Failed to install package:", pkg, "Error:", e$message))
        return(FALSE)
      })
  } else {
    library(pkg, character.only = TRUE)
    return(TRUE)
  }
})

if (!packages_loaded["data.table"]) {
  stop("Critical package 'data.table' could not be loaded. Aborting.")
}
message("--- Package Loading Complete ---")

csv_folder_path <- "/Users/alex/Desktop/ECO340-Project" # USER: MODIFY THIS PATH
tryCatch({
    setwd(csv_folder_path)
    message(paste("Working directory set to:", getwd()))
  }, warning = function(w){
    stop("Warning setting working directory: ", w$message, ". Check path: ", csv_folder_path)
  }, error = function(e) {
    stop("Error setting working directory: ", e$message, ". Check path: ", csv_folder_path)
  }
)

message("--- Starting Data Loading ---")
message("Listing all CSV files...")
all_csv_files <- list.files(path = getwd(), pattern = "\\.csv$", full.names = TRUE, ignore.case = TRUE)

if (length(all_csv_files) == 0) {
  stop("No CSV files found in directory: ", getwd())
} else {
   message(paste("Found", length(all_csv_files), "total CSV files."))
}

filename_pattern <- "^pub(01|06)\\d{2}\\.csv$"
message(paste("Filtering for files matching pattern:", filename_pattern))

all_basenames <- basename(all_csv_files)
matching_indices <- grepl(filename_pattern, all_basenames, ignore.case = TRUE)
csv_files <- all_csv_files[matching_indices]

if (length(csv_files) == 0) {
  stop("No CSV files matching the pattern '", filename_pattern, "' were found in ", getwd())
} else {
  message(paste("Found", length(csv_files), "files matching the pattern. Processing..."))
  message("Files to be processed:")
  print(basename(csv_files))

  LFS_data <- NULL
  for (i in seq_along(csv_files)) {
    file_path <- csv_files[i]
    message(paste("Processing file", i, "of", length(csv_files), ":", basename(file_path)))

    current_data <- tryCatch({
        data.table::fread(file_path, showProgress = FALSE)
      }, error = function(e) {
        message(paste("Error reading file:", basename(file_path), "-", e$message))
        return(NULL)
      })

    if (!is.null(current_data) && nrow(current_data) > 0) {
      if (is.null(LFS_data)) {
        LFS_data <- current_data
      } else {
        LFS_data <- data.table::rbindlist(list(LFS_data, current_data), use.names = TRUE, fill = TRUE)
      }
    } else if (!is.null(current_data) && nrow(current_data) == 0) {
        message(paste("Skipping empty file:", basename(file_path)))
    }
    gc()
  }

  gc()
  message("--- Data Loading Loop Finished ---")

  if (!is.null(LFS_data) && data.table::is.data.table(LFS_data) && nrow(LFS_data) > 0) {
    message("Data loaded and combined successfully.")
    message("Dimensions of combined data (rows, columns):")
    print(dim(LFS_data))
    message("Memory usage:")
    print(object.size(LFS_data), units = "auto")
  } else {
      message("Warning: Final combined data is empty or invalid. Check source files and logs.")
      if (!is.null(LFS_data)) print(str(LFS_data))
  }
}

message("--- Script Finished ---")


```

## Introduction

Over the years, Canada's Gini coefficient with respect to income has fluctuated over the years. This analysis endeavours to show if there have been any meaningful trends in inequities within different age and gender groupings of the Canadian population. The first part of this analysis will be concerned with the Gini coeffients for each age group plotted over the past 24 years, using data from the Canadian LFS from March 2000 through to March 2024. The second part of the analysis will be looking at trends in unemployment and labour force participation over this same period of time for each group.

As this data is not like the NLSY in that it follows the same people over the years, this will not be a generational/cohort study, ie the age brackets will remain fixed and over the years the cohorts will move through them. As well, as this analysis is using the publicly available version of the LFS, data is not provided on race or certain other potentially identifying characteristics.

##Data Cleaning & Variable Creation

To make analysis easier, some new variables will be created to make an easier time of things. Variables which give multiple categories in numeric form, for example province codes, will be converted to binary dummy variables. The variables that will be converted in this fashion are: gender, re-coding from 1 & 2 to 0 & 1; union status, breaking the "union member" "union contract" and "neither" into two dummy variables for "union member" and "union contract"; the aformentioned provinces, creating 9 dummy variables with Ontario being the reference province; education, creating two sets of dummy variables one being a culmulative set (ie each person only has one dummy variable that's in the on position) and the other being individual (ie someone with a college education has college turned on as well as every switch before); age, recoded to be an approximate age using both AGE12 and AGE6 data to maximize precision; LFSSTAT, recoded from a 1, 2, 3, 4 to a "in labour force", "is employed", and "absent from work" dummies; immigrant similarly into two dummies, one for recent immigrant one for established immigrant; year and month data into months from start of data-set (I changed directions so didn't really use this). Also creating a logarithmic version of hourly earnings so that regressions can be easily interpereted.

```{r echo=FALSE}
#For simplicity we will break down the variable LFSSTAT into an Employed binary variable and an in labour force variable. Let's label these is_employed and in_lf -- using these we can more easily down the line calculated changing unemployment rates and labour force participation rates.
#Gender in the LFS datafile is expressed as either a 1 for men and 2 for women, stored as GENDER. We will create a new variable, woman, which will be a binary variable -- 0 for men, 1 for women.
#as opposed to using province codes, I have created new binary variables for the 9 provinces which aren't Ontario. Because I am biased, I will be using Ontario as the default.
#similar to the step for dis-aggregation of the LFSSTAT variable, will decompose the union variable, which gives a 1 for being a union member, a 2 for being covered by union contract, and a 3 for being not covered (an 8 for not being applicable, but dwai bc we won't run a regression on them), into a binary for being a union member (ie 1 only) and a binary for being covered by union contract (1 and 2). we will call these union_member and union_contract (outstanding creativity there w the names)
#same process now for IMMIG, immigration status. going to create variables newer_immigrant for immigrants here for 10 or less years, older_immigrant for those who have been here for more than 10 years.
#education will be broken into culmulative dummies, NOTE TO SELF NEED TO REMEMBER THERE IS SIG COLLINEARITY MUST ACCOUNT FOR, as well as a set of exclusive categories for regression purposes :)
# Assuming LFS_data is your data frame
# Convert LFS_data to a data.table object *in place*
# This is crucial for memory efficiency.
setDT(LFS_data)

#Log Hourly Earnings
LFS_data[, ln_income := log(HRLYEARN)]

# --- Create binary indicator variables using := ---
# The `:=` operator modifies the data.table in place, avoiding copies.

# Employment and Labour Force Status
LFS_data[, is_employed := fifelse(LFSSTAT == 1, 1, 0)]
LFS_data[, in_lf := fifelse(LFSSTAT %in% c(1, 2), 1, 0)]

# Sex
LFS_data[, woman := fifelse(SEX == 2 | GENDER == 2, 1L, 0L, na = 0L)]

# Age above 15, proxy for how many years as an adult either receiving education or in the workforce -- while I would like to have exact years of age and years of education and then subtract one from the other - 5 to account for the difference, there is more limited data here so a rough age approximation by taking the midpoint of the age brackets will be done -- yeah I COULD look at Canada's population pyramid and make a weighted estimate using that but then again stats Canada does there own controls and blah blah
# Below takes the midpoint of each age range and then docks 15 from it
LFS_data[, rough_age_dock15 := fcase(
    AGE_12 == 1,  2,
    AGE_12 == 2,  7,
    AGE_12 == 3,  12,
    AGE_12 == 4,  17,
    AGE_12 == 5,  22,
    AGE_12 == 6,  27,
    AGE_12 == 7,  32,
    AGE_12 == 8,  37,
    AGE_12 == 9,  42,
    AGE_12 == 10, 47,
    AGE_12 == 11, 52,
    AGE_12 == 12, 57,
    default = NA_real_
)]

# Province Indicators (using fifelse for speed)
# Note: You could combine these into fewer lines if preferred,
# but separating them is clear.
LFS_data[, prov_nl := fifelse(PROV == 10, 1, 0)]
LFS_data[, prov_pe := fifelse(PROV == 11, 1, 0)]
LFS_data[, prov_ns := fifelse(PROV == 12, 1, 0)]
LFS_data[, prov_nb := fifelse(PROV == 13, 1, 0)]
LFS_data[, prov_qc := fifelse(PROV == 24, 1, 0)]
LFS_data[, prov_mb := fifelse(PROV == 46, 1, 0)]
LFS_data[, prov_sk := fifelse(PROV == 47, 1, 0)]
LFS_data[, prov_ab := fifelse(PROV == 48, 1, 0)]
LFS_data[, prov_bc := fifelse(PROV == 59, 1, 0)]
# For ease of making regressions later
province_dummies <- c("prov_nl", "prov_pe", "prov_ns", "prov_nb", "prov_qc", "prov_mb", "prov_sk", "prov_ab", "prov_bc")

# Union Status
LFS_data[, union_member := fifelse(UNION == 1, 1, 0)]
LFS_data[, union_contract := fifelse(UNION %in% c(1, 2), 1, 0)]

# Immigration Status
LFS_data[, newer_immigrant := fifelse(IMMIG == 1, 1, 0)]
LFS_data[, older_immigrant := fifelse(IMMIG == 2, 1, 0)]

# Cumulative Education Indicators
LFS_data[, some_hs_plus := fifelse(EDUC >= 1, 1, 0)]
LFS_data[, hs_grad_plus := fifelse(EDUC >= 2, 1, 0)]
LFS_data[, some_ps_plus := fifelse(EDUC >= 3, 1, 0)]
LFS_data[, ps_cert_plus := fifelse(EDUC >= 4, 1, 0)]
LFS_data[, ba_plus := fifelse(EDUC >= 5, 1, 0)]
LFS_data[, above_ba_plus := fifelse(EDUC == 6, 1, 0)]

# Exclusive Education Indicators
LFS_data[, zero_to_eight := fifelse(EDUC == 0, 1, 0)]
LFS_data[, some_hs := fifelse(EDUC == 1, 1, 0)]
LFS_data[, some_ps := fifelse(EDUC == 3, 1, 0)]
LFS_data[, ps_cert := fifelse(EDUC == 4, 1, 0)]
LFS_data[, ba := fifelse(EDUC == 5, 1, 0)]
LFS_data[, above_ba := fifelse(EDUC == 6, 1, 0)]
#for ease of regressions later
education_dummies <- c("zero_to_eight", "some_hs", "some_ps", "ps_cert", "ba", "above_ba")

LFS_data[, naics_ForestryLogSupport := fifelse(NAICS_21 == 2, 1, 0)]
LFS_data[, naics_FishingHuntTrap := fifelse(NAICS_21 == 3, 1, 0)]
LFS_data[, naics_MiningOilGasExtract := fifelse(NAICS_21 == 4, 1, 0)]
LFS_data[, naics_Utilities := fifelse(NAICS_21 == 5, 1, 0)]
LFS_data[, naics_Construction := fifelse(NAICS_21 == 6, 1, 0)]
LFS_data[, naics_MfgDurable := fifelse(NAICS_21 == 7, 1, 0)] # Manufacturing - durable goods
LFS_data[, naics_MfgNonDurable := fifelse(NAICS_21 == 8, 1, 0)] # Manufacturing - non-durable goods
LFS_data[, naics_Wholesale := fifelse(NAICS_21 == 9, 1, 0)]
LFS_data[, naics_Retail := fifelse(NAICS_21 == 10, 1, 0)]
LFS_data[, naics_TransportWarehousing := fifelse(NAICS_21 == 11, 1, 0)]
LFS_data[, naics_FinanceInsurance := fifelse(NAICS_21 == 12, 1, 0)]
LFS_data[, naics_RealEstateRentalLeasing := fifelse(NAICS_21 == 13, 1, 0)]
LFS_data[, naics_ProfSciTech := fifelse(NAICS_21 == 14, 1, 0)] # Professional, scientific and technical services
LFS_data[, naics_BusinessBuildingSupport := fifelse(NAICS_21 == 15, 1, 0)] # Business, building and other support services
LFS_data[, naics_Educational := fifelse(NAICS_21 == 16, 1, 0)]
LFS_data[, naics_HealthSocialAssist := fifelse(NAICS_21 == 17, 1, 0)] # Health care and social assistance
LFS_data[, naics_InfoCultRec := fifelse(NAICS_21 == 18, 1, 0)] # Information, culture and recreation
LFS_data[, naics_AccomFood := fifelse(NAICS_21 == 19, 1, 0)]
LFS_data[, naics_OtherServices := fifelse(NAICS_21 == 20, 1, 0)]
LFS_data[, naics_PublicAdmin := fifelse(NAICS_21 == 21, 1, 0)]

# Compile the names of the created NAICS dummy variables into a vector (updated names)
naics_dummies <- c(
  "naics_ForestryLogSupport", "naics_FishingHuntTrap", "naics_MiningOilGasExtract", "naics_Utilities",
  "naics_Construction", "naics_MfgDurable", "naics_MfgNonDurable", "naics_Wholesale",
  "naics_Retail", "naics_TransportWarehousing", "naics_FinanceInsurance", "naics_RealEstateRentalLeasing",
  "naics_ProfSciTech", "naics_BusinessBuildingSupport", "naics_Educational", "naics_HealthSocialAssist",
  "naics_InfoCultRec", "naics_AccomFood", "naics_OtherServices", "naics_PublicAdmin"
)



# --- Create years_from_start ---
# Assuming LFS_data exists and has SURVYEAR, SURVMNTH
# Assuming the start month is January 2006 (index 1)
LFS_data[, years_from_start := (SURVYEAR - 2006) + ifelse(SURVMNTH == 6, 0.5, 0)]

# --- Create Dummy Variables and Store Names ---

# --- Method 1: Capture names DURING creation ---
# Initialize an empty character vector to store the names
yfs_dummy_names <- character(0) # Use a descriptive name for the vector

# Get unique non-NA values from the source column
unique_vals <- sort(unique(na.omit(LFS_data$years_from_start)))

# Check if there are unique values to process
if (length(unique_vals) > 0) {
  print(paste("Starting loop to create", length(unique_vals), "dummy columns..."))

  for (val in unique_vals) {
    # Create a valid column name (e.g., yfs_0, yfs_0_5)
    new_col_name <- paste0("yfs_", gsub("[^a-zA-Z0-9_]", "_", as.character(val)))

    # *** Correctly append the generated name to the VECTOR ***
    yfs_dummy_names <- c(yfs_dummy_names, new_col_name)

    # Create the dummy column in LFS_data using fifelse
    # This modifies LFS_data in place
    LFS_data[, (new_col_name) := fifelse(years_from_start == val, 1L, 0L, na = 0L)]
  }
  print("Finished creating dummy columns.")
  print("Dummy names captured in the vector 'yfs_dummy_names'")

  # You can view the captured names:
  # print(yfs_dummy_names)

} else {
  print("No unique non-NA values found in 'years_from_start' to create dummies for.")
}



print("Finished creating dummy columns.")
print("Dummy names captured in 'yfs_dummies'")
# print(yfs_dummies) # View the captured names


# --- Creating 'province_name' column from PROV code ---
# Doing this to make creating subsets easier in next block
LFS_data[, province_name := fcase(
    PROV == 10, "Newfoundland and Labrador",
    PROV == 11, "Prince Edward Island",
    PROV == 12, "Nova Scotia",
    PROV == 13, "New Brunswick",
    PROV == 24, "Quebec",
    PROV == 35, "Ontario",
    PROV == 46, "Manitoba",
    PROV == 47, "Saskatchewan",
    PROV == 48, "Alberta",
    PROV == 59, "British Columbia"
)]

# --- Creating Male/Female named variable --- 
# See above
LFS_data[, named_sex := fcase(
    woman == 0, "Male",
    woman == 1, "Female"
)]

# --- Creating Age Cohort named variable --- 
# See above
# The age groups are going to be 15-19, 20-34, 35-54, and then 55+ for simplicity
LFS_data[, age_group := fcase(
    AGE_12 == 1, "15_19",
    AGE_12 %in% c(2, 3, 4), "20_34",
    AGE_12 %in% c(5, 6, 7, 8), "35_54",
    AGE_12 %in% c(9, 10, 11, 12), "55+",
    default = NA_character_
)]

gc()
```

## Building a Model of Hourly Wages

This section outlines the construction of a regression model used to determine the impact of key variables and controls on hourly earnings. As this model will be estimated by applying it to multiple waves of LFS data, year dummy variables will be included to account for time fixed-effects/year specific events. This model will be restricted to employed Canadians -- variations in unemployment and labour force participation will be covered later on.

The dependent variable in these regressions will be the natural logarithm of hourly earnings, for easier interpretation of coefficients. The "base" or intercept observation, with respect to dummy variables, is a male high school graduate from Ontario who is not in a union, nor is covered by a union contract, who is not an immigrant, and who works in the Agriculture sector.

``` {r echo=FALSE}
#Note: make sure to overwrite regression outputs for memory efficiency, then print/store coefficients only -- cannot sustain full models.

# Store the models for comparison
models <- list()

# Model 1: Basic with gender and year effects only
dummy_formula_part <- paste(yfs_dummy_names, collapse = " + ")
regression_formula <- as.formula(paste("ln_income ~ woman +", dummy_formula_part))
# Fit the linear model using lm function
models[[1]] <- lm(regression_formula, data = LFS_data)

# Model 2: Adding province effects
dummy_formula_part1 <- paste(yfs_dummy_names, collapse = " + ")
dummy_formula_part2 <- paste(province_dummies, collapse = " + ")
# Construct the formula string dynamically
regression_formula <- as.formula(paste("ln_income ~ woman + ", dummy_formula_part1, " + ", dummy_formula_part2))
# Fit the linear model
models[[2]] <- lm(regression_formula, data = LFS_data)

# Model 3: Adding education effects
dummy_formula_part1 <- paste(yfs_dummy_names, collapse = " + ")
dummy_formula_part2 <- paste(province_dummies, collapse = " + ")
dummy_formula_part3 <- paste(education_dummies, collapse = " + ")
# Construct the formula string dynamically
regression_formula <- as.formula(paste("ln_income ~ woman + ", dummy_formula_part1, " + ", dummy_formula_part2, " + ", dummy_formula_part3))
# Fit the linear model
models[[3]] <- lm(regression_formula, data = LFS_data)

# Model 4: Adding age polynomial
dummy_formula_part1 <- paste(yfs_dummy_names, collapse = " + ")
dummy_formula_part2 <- paste(province_dummies, collapse = " + ")
dummy_formula_part3 <- paste(education_dummies, collapse = " + ")
# Construct the formula string including a polynomial term for age
regression_formula <- as.formula(paste("ln_income ~ woman + poly(rough_age_dock15, 2, raw=TRUE) + ",
                                     dummy_formula_part1, " + ", dummy_formula_part2, " + ", dummy_formula_part3))
# Fit the linear model
models[[4]] <- lm(regression_formula, data = LFS_data)

# Model 5: Full model with union, immigration, and industry sectors
dummy_formula_part1 <- paste(yfs_dummy_names, collapse = " + ")
dummy_formula_part2 <- paste(province_dummies, collapse = " + ")
dummy_formula_part3 <- paste(education_dummies, collapse = " + ")
dummy_formula_part4 <- paste(naics_dummies, collapse = " + ")
# Construct the formula string for the full model
regression_formula <- as.formula(paste("ln_income ~ woman + poly(rough_age_dock15, 2, raw=TRUE) + newer_immigrant + older_immigrant + union_contract + ",
                                     dummy_formula_part1, " + ", dummy_formula_part2, " + ", dummy_formula_part3, " + ", dummy_formula_part4))
# Fit the linear model
models[[5]] <- lm(regression_formula, data = LFS_data)

# Define model names to include all 5 models for the summary table
model_names <- c("Gender Only", "Add Provinces", "Add Education", "Add Age", "Full Model")

# *** Add names to the models list ***
# This assigns the defined names to the list elements,
# which modelsummary uses as column headers.
names(models) <- model_names

# Create a custom coefficient map for renaming variables in the output table
# This map improves readability of the regression table
coef_map <- c(
  "woman" = "Female",
  "poly(rough_age_dock15, 2, raw = TRUE)1" = "Age",
  "poly(rough_age_dock15, 2, raw = TRUE)2" = "Age²",
  "newer_immigrant" = "Recent Immigrant (≤10yrs)",
  "older_immigrant" = "Established Immigrant (>10yrs)",
  "union_contract" = "Union Coverage",

  # Keep only key education variables for brevity
  "zero_to_eight" = "Education: 0-8 years",
  "ba" = "Education: Bachelor's degree",
  "above_ba" = "Education: Above bachelor's",

  # Only selected provinces for brevity
  "prov_qc" = "Quebec",
  "prov_ab" = "Alberta",
  "prov_bc" = "British Columbia",

  # Three selected industry sectors for brevity
  "naics_MiningOilGasExtract" = "Industry: Mining & Oil Extraction",
  "naics_FinanceInsurance" = "Industry: Finance & Insurance",
  "naics_HealthSocialAssist" = "Industry: Healthcare & Social Assist."
)

# Create table with modelsummary and set model names explicitly
# modelsummary generates a summary table of regression models
modelsummary(models, # Pass the NAMED list of models
             title = "Log Wage Regressions with Different Specifications",
             stars = TRUE, # Add significance stars
             coef_map = coef_map, # Apply the custom coefficient names
             gof_map = c("nobs", "r.squared", "adj.r.squared"), # Select goodness-of-fit statistics
             notes = "Year fixed effects included in all models. Ontario is the reference province. Full range of dummy variables not shown for ease of viewing", # Add notes to the table
             output = "html") # Specify HTML output format

```

## Group Creation

To make this next part easier, subsets of our LFS_data dataframe can be created. Think of a fork in the road type structure -- the order will be as follows: fork by province, fork by gender, then fork by age. Later on when retrieving our subsetted data, the syntax to retrieve will be: subsetted_data\$PROVINCE\$SEX\$AGE

```{r}
gc()
rm(models) #memory constrained, need to drop prior bulky derived data
#create a list of province based subsets
subsetted_data <- split(LFS_data, LFS_data$province_name) #to retrieve subset, province_subset_list$PROVINCE

#now inception style we split again based on sex
subsetted_data <- lapply(subsetted_data, function(subset_prov) {
  split(subset_prov, subset_prov$named_sex) #to retrieve subset, subsetted_data$PROVINCE$SEX
})

#third layer, we now divide the aforementioned subsets into out four age categories
subsetted_data <- lapply(subsetted_data, function(subset_prov) {
  lapply(subset_prov, function(subset_prov_sex){
    split(subset_prov_sex, subset_prov_sex$age_group) #to retrieve subset, province_sex_age_subset_list$PROVINCE$SEX$AGEGROUP
  })
})

```

## Limitations of the Data

As this paper is using the publicly available LFS data, certain information is excluded -- thanks to this there are no observations for the Territories, nor is there any access to ethnicity/race variables. Finally, the standard scope of the LFS applies: full time members of the armed forces, people living on reserves, institutionalized persons, and those under the age of 15 are excluded.

## AI Disclosures:

Through the course of this project, I utilized Gemini. Unlike ChatGPT Gemini does not provide an easy way to share the whole chat, so I've included some sample queries and responses and wrote a quick blurb explaining how the tools were used.

set-up: <https://g.co/gemini/share/085ca32bc3d4>, <https://g.co/gemini/share/429555247dcd> This was crucial, only noticed that I was doing a very inefficient way to merge the csv files when I swapped from working on my desktop with 64Gb of RAM to my MacBook with 16Gb lol. I also used it to make a little loop function for loading the packages, just to make it easier to add new packages I want to use as I go through the project. I actually wound up just taking sample observations twice a year -- I wanted to use all 13+ million observations over the period, but having to reprogram/adjust how I did so many things to stop hitting memory limits just got tiresome.

general writing: <https://chatgpt.com/share/67eebc76-c988-8010-801d-fbc4827958f2> I chiefly use this as a tool to help catch where I have been unprofessional/super casual in my writing -- I paste a section of my text into the LLM and then I ask it to point out areas for improvement, then I go back and edit it, and so on and so forth in a recursive process till I'm happy with the result/the AI is just splitting hairs. The key improvement here is that I bug friends and family less with requests to read over things.

general code and data-set questions: <https://g.co/gemini/share/cc0925dc8fb8>, <https://g.co/gemini/share/6866b019b792> <https://g.co/gemini/share/898697efa94a>

Main workflow for code corrections is to boot up a chat with Gemini, say hey I'm working in R in a markdown file, can you correct this chunk of code for me? I tried Cursor (an AI integrated IDE) but found that it just made it too tempting/easy to vibecode the whole thing and not know what each part of my code even does. Going chunk by chunk I found made it so that AI was helping my weaker knowledge of R syntax, but that I was still responsible for the overall structure of my code and for deciding how to structure, say, my subgroups and analysis. (The question is "Hey how can I made subsets by a variable" not "Here's the file: create subsets by what variables you think are best to divvy up the population)
